---
title: "Clean Personnel"
format: html
---

# Setup

```{r}
library(tidyverse)
library(glue)
library(janitor)
library(readxl)
library(writexl)
library(arrow)

options(scipen = 12, digits = 12)

path_data <- Sys.getenv("PATH_DATA")
year_fiscal <- Sys.getenv("YEAR_FISCAL")
```

# Import

Ledger

```{r}
ledger <-
  read_parquet(glue("{path_data}/ledger.parquet")) %>%
  filter(
    account_category == "700000A",
    account_id != "7012999"
  ) %>%
  filter(ending_balance != 0.00)

ledger_rollup <-
  ledger %>%
  summarize(expense_ledger = sum(ending_balance), .by = c(
    account_id, account_name,
  )) %>%
  arrange(account_id) %>%
  filter(expense_ledger != 0.00)
```

Personnel

```{r}
data_485_raw <-
  read_parquet(glue("{path_data}/personnel_{year_fiscal}.parquet")) %>%
  replace_na(list(project_id = "000000", award_id = "0000000")) %>%
  select(!c(row_485, account_raw, account_pref_2, account_pref_4))

data_485_raw %>% nrow()
```

# Tidy

Initial cleaning

```{r}
program_names <-
  ledger %>%
  distinct(program_id, program_name)

costcenter_names <-
  ledger %>%
  distinct(costcenter_id, costcenter_name)

project_names <-
  ledger %>%
  distinct(project_id, project_name)

award_names <-
  ledger %>%
  distinct(award_id, award_name)

position_titles <-
  data_485_raw %>%
  distinct(position_id, position_title, pay_end_dt) %>%
  slice_max(order_by = pay_end_dt, n = 1, by = position_id) %>%
  select(position_id, position_title)

employee_names <-
  data_485_raw %>%
  distinct(employee_id, employee_name, pay_end_dt) %>%
  slice_max(order_by = pay_end_dt, n = 1, by = employee_id) %>%
  select(employee_id, employee_name)
```

```{r}
data_485 <-
  data_485_raw %>%
  select(!c(
    program_name,
    costcenter_name,
    project_name,
    award_name,
    position_title,
    employee_name
  )) %>%
  left_join(program_names, by = join_by(program_id)) %>%
  left_join(costcenter_names, by = join_by(costcenter_id)) %>%
  left_join(project_names, by = join_by(project_id)) %>%
  left_join(award_names, by = join_by(award_id)) %>%
  left_join(position_titles, by = join_by(position_id)) %>%
  left_join(employee_names, by = join_by(employee_id))
```

Find correct programs for payroll default records.

```{r}
has_program_position <-
  data_485 %>%
  filter(!is.na(program_name)) %>%
  slice_max(
    order_by = pay_end_dt, n = 1, by = position_id, with_ties = FALSE
  ) %>%
  select(position_id, program_id, program_name)

has_program_employee <-
  data_485 %>%
  filter(!is.na(program_name)) %>%
  slice_max(
    order_by = pay_end_dt, n = 1, by = employee_id, with_ties = FALSE
  ) %>%
  select(employee_id, program_id, program_name)

program_foundby_position <-
  data_485 %>%
  filter(is.na(program_name)) %>%
  distinct(position_id) %>%
  filter(position_id %in% has_program_position$position_id) %>%
  left_join(has_program_position, by = join_by(position_id)) %>%
  arrange(program_name, position_id)

program_foundby_employee <-
  data_485 %>%
  filter(is.na(program_name)) %>%
  filter(!(position_id %in% program_foundby_position$position_id)) %>%
  distinct(employee_id) %>%
  filter(employee_id %in% has_program_employee$employee_id) %>%
  left_join(has_program_employee, by = join_by(employee_id)) %>%
  arrange(program_name, employee_id)
```

Find correct cost centers for payroll default records.

```{r}
has_costcenter_position <-
  data_485 %>%
  filter(program_id != "150011") %>%
  slice_max(
    order_by = pay_end_dt, n = 1, by = position_id, with_ties = FALSE
  ) %>%
  select(position_id, program_id, program_name, costcenter_id, costcenter_name)

has_costcenter_employee <-
  data_485 %>%
  filter(program_id != "150011") %>%
  slice_max(
    order_by = pay_end_dt, n = 1, by = employee_id, with_ties = FALSE
  ) %>%
  select(employee_id, program_id, program_name, costcenter_id, costcenter_name)

costcenter_foundby_position <-
  data_485 %>%
  filter(program_id == "150011") %>%
  distinct(position_id) %>%
  filter(position_id %in% has_costcenter_position$position_id) %>%
  left_join(has_costcenter_position, by = join_by(position_id))

costcenter_foundby_employee <-
  data_485 %>%
  filter(program_id == "150011") %>%
  filter(!(position_id %in% costcenter_foundby_position$position_id)) %>%
  distinct(employee_id) %>%
  filter(employee_id %in% has_costcenter_employee$employee_id) %>%
  left_join(has_costcenter_employee, by = join_by(employee_id))
```

```{r}
data_485_deffound_position <-
  data_485 %>%
  filter(program_id == "150011") %>%
  filter(position_id %in% costcenter_foundby_position$position_id) %>%
  select(!c(program_id, program_name, costcenter_id, costcenter_name)) %>%
  left_join(costcenter_foundby_position, by = join_by(position_id))

data_485_deffound_employee <-
  data_485 %>%
  filter(program_id == "150011") %>%
  filter(!(position_id %in% costcenter_foundby_position$position_id)) %>%
  filter(employee_id %in% costcenter_foundby_employee$employee_id) %>%
  select(!c(program_id, program_name, costcenter_id, costcenter_name)) %>%
  left_join(costcenter_foundby_employee, by = join_by(employee_id))

data_485_defnotfound <-
  data_485 %>%
  filter(program_id == "150011") %>%
  filter(!(position_id %in% costcenter_foundby_position$position_id)) %>%
  filter(!(employee_id %in% costcenter_foundby_employee$employee_id))

data_485 <-
  data_485 %>%
  filter(program_id != "150011") %>%
  bind_rows(
    data_485_deffound_position,
    data_485_deffound_employee,
    data_485_defnotfound
  )

data_485 %>% nrow()
```

Diagnostics

```{r}
ledger %>%
  pull(ending_balance) %>%
  sum()

data_485 %>%
  filter(!(program_id %in% ledger$program_id)) %>%
  pull(amount) %>%
  sum()

data_485 %>%
  filter(!(costcenter_id %in% ledger$costcenter_id)) %>%
  pull(amount) %>%
  sum()

data_485 %>%
  filter(!(account_id %in% ledger$account_id)) %>%
  pull(amount) %>%
  sum()

data_485 %>%
  filter(is.na(position_title)) %>%
  pull(amount) %>%
  sum()

data_485 %>%
  filter(is.na(employee_name)) %>%
  pull(amount) %>%
  sum()
```

Filter out programs, cost centers, and accounts missing from ledger.
Filter out records missing position title or employee name.

```{r}
data_485 <-
  data_485 %>%
  filter(
    program_id %in% ledger$program_id,
    costcenter_id %in% ledger$costcenter_id,
    account_id %in% ledger$account_id,
  ) %>%
  filter(
    !is.na(position_title),
    !is.na(employee_name),
  )
```

Compare accounts in ledger to accounts in 485.

Find ledger accounts missing from 485.
For these accounts, additional rows will be imputed into the 485.

Find accounts present in ledger and in 485.
For these accounts, a multiplier will calculated per account so that personnel 
spending in the 485 equals personnel spending in the ledger.

```{r}
personnel_rollup <-
  data_485 %>%
  summarize(expense_485 = sum(amount), .by = c(
    account_id
  )) %>%
  arrange(account_id) %>%
  filter(expense_485 != 0.00)

personnel_comparison <-
  ledger_rollup %>%
  full_join(personnel_rollup, by = join_by(
    account_id
  ))

accounts_missing_485 <-
  personnel_comparison %>%
  filter(is.na(expense_485)) %>%
  select(account_id, expense_ledger)

multipliers <-
  personnel_comparison %>%
  filter(!is.na(expense_485)) %>%
  mutate(multiplier = expense_ledger / expense_485) %>%
  select(account_id, multiplier)
```

Impute

```{r}
attributes_imputable <-
  ledger %>%
  filter(account_id %in% accounts_missing_485$account_id) %>%
  distinct(
    fund_id, program_id, costcenter_id,
    account_id
  ) %>%
  mutate(is_imputed = TRUE) %>%
  pivot_wider(
    names_from = account_id, names_sort = TRUE,
    values_from = is_imputed
  )

rows_need_impute <-
  data_485 %>%
  left_join(attributes_imputable, by = join_by(
    fund_id, program_id, costcenter_id,
  )) %>%
  filter(if_any(`7013015`:`7014021`)) %>%
  select(!c(amount, account_id)) %>%
  pivot_longer(
    cols = c(`7013015`:`7014021`),
    names_to = "account_id", values_to = "is_imputed", values_drop_na = TRUE
  ) %>%
  select(!is_imputed)

amounts_impute <-
  rows_need_impute %>%
  count(account_id) %>%
  left_join(accounts_missing_485, by = join_by(account_id)) %>%
  mutate(amount = expense_ledger / n) %>%
  select(account_id, amount) %>%
  arrange(account_id)

rows_imputed <-
  rows_need_impute %>%
  left_join(amounts_impute, by = join_by(account_id))
```

```{r}
data_485_corrected <-
  data_485 %>%
  left_join(multipliers, by = join_by(account_id)) %>%
  mutate(amount = amount * multiplier) %>%
  select(!multiplier) %>%
  bind_rows(rows_imputed)

data_485_corrected %>%
  pull(amount) %>%
  sum()

ledger %>%
  pull(ending_balance) %>%
  sum()
```

# Transform

```{r}
personnel_clean <-
  data_485_corrected %>%
  mutate(amt_salary = case_when(
    str_starts(account_id, "7011") ~ amount,
    str_starts(account_id, "7012") ~ amount,
    str_starts(account_id, "7013") ~ amount,
    str_starts(account_id, "7015") ~ amount,
    str_starts(account_id, "7016") ~ amount,
    .default = 0.00
  )) %>%
  mutate(amt_benefits = case_when(
    str_starts(account_id, "7014") ~ amount,
    .default = 0.00
  )) %>%
  mutate(amt_terminal = case_when(
    account_id == "7013007" ~ amount,
    .default = 0.00
  )) %>%
  summarize(
    amt_salary = sum(amt_salary),
    amt_benefits = sum(amt_benefits),
    amt_terminal = sum(amt_terminal),
    .by = c(
      costcenter_id, costcenter_name,
      program_id, program_name,
      position_id, position_title,
      employee_id, employee_name
    )
  ) %>%
  arrange(
    costcenter_name,
    program_name,
    position_title,
    employee_name
  )
```

# Output

```{r}
personnel_clean %>%
  write_parquet(glue("{path_data}/personnel_clean.parquet"))

personnel_clean %>%
  write_csv(glue("{path_data}/personnel_clean.csv"), na = "")
```

```{r}
Sys.time()
```
