---
title: "Clean Personnel"
format: html
---

# Setup

```{r}
library(tidyverse)
library(glue)
library(janitor)
library(readxl)
library(writexl)
library(arrow)

options(scipen = 12, digits = 12)

path_data <- Sys.getenv("PATH_DATA")
year_fiscal <- Sys.getenv("YEAR_FISCAL")

.data <- rlang::.data
```

# Import

## Categorization

```{r}
costcenters_school <-
  read_excel(glue("{path_data}/manual/costcenters_categorization.xlsx")) |>
  select(costcenter_id, is_school)

costcenters_categorization <-
  read_excel(glue("{path_data}/manual/costcenters_categorization.xlsx")) |>
  select(costcenter_id, is_school, is_all_direct, is_all_indirect, is_mixed)

positions_direct <-
  read_excel(glue("{path_data}/manual/positions_direct.xlsx")) |>
  pull(position_title) |>
  str_flatten(collapse = "|")
```

## Ledger

Filter for personnel.
Filter out non-budgetary.

```{r}
ledger_full <-
  read_parquet(glue("{path_data}/working/ledger.parquet")) |>
  left_join(costcenters_school, by = join_by(costcenter_id)) |>
  mutate(
    costcenter_id = case_when(
      is_school ~ "SCHOOL",
      costcenter_id %in% c("10001", "10002", "10003") ~ "OCFO",
      costcenter_id %in% c(
        "40080", "40084", "40228", "40229", "40242"
      ) ~ "OTHER DIRECT",
      costcenter_id %in% c("00000", "40223") ~ "OTHER INDIRECT",
      .default = costcenter_id
    ),
    costcenter_name = case_when(
      costcenter_id == "SCHOOL" ~ "SCHOOL",
      costcenter_id == "OCFO" ~ "OFFICE OF THE CHIEF FINANCIAL OFFICER",
      costcenter_id == "OTHER DIRECT" ~ "OTHER DIRECT",
      costcenter_id == "OTHER INDIRECT" ~ "OTHER INDIRECT",
      .default = costcenter_name
    ),
  )

ledger_personnel <-
  ledger_full |>
  filter(
    account_category == "700000A", # personnel
    account_id != "7012999", # non-budgetary
  ) |>
  mutate(account_4 = str_extract(account_id, "^701[1-9]"))

costcenters_nonzero <-
  ledger_personnel |>
  summarize(
    ending_balance = sum(ending_balance),
    .by = costcenter_id
  ) |>
  filter(ending_balance != 0.00)

costcenters_negative <-
  costcenters_nonzero |>
  filter(ending_balance < 0.00)

accounts_4_nonzero <-
  ledger_personnel |>
  summarize(ending_balance = sum(ending_balance), .by = account_4) |>
  filter(ending_balance != 0.00)
```

## Personnel

```{r}
personnel <-
  read_parquet(glue("{path_data}/raw/personnel_{year_fiscal}.parquet")) |>
  left_join(costcenters_school, by = join_by(costcenter_id)) |>
  mutate(
    costcenter_id = case_when(
      is_school ~ "SCHOOL",
      costcenter_id %in% c("10001", "10002", "10003") ~ "OCFO",
      costcenter_id %in% c(
        "40080", "40084", "40228", "40229", "40242"
      ) ~ "OTHER DIRECT",
      costcenter_id %in% c("00000", "40223") ~ "OTHER INDIRECT",
      .default = costcenter_id
    ),
    costcenter_name = case_when(
      costcenter_id == "SCHOOL" ~ "SCHOOL",
      costcenter_id == "OCFO" ~ "OFFICE OF THE CHIEF FINANCIAL OFFICER",
      costcenter_id == "OTHER DIRECT" ~ "OTHER DIRECT",
      costcenter_id == "OTHER INDIRECT" ~ "OTHER INDIRECT",
      .default = costcenter_name
    ),
  ) |>
  mutate(
    costcenter_id = if_else(
      program_id == "150011", "PAYROLL DEFAULT", costcenter_id
    ),
    costcenter_name = if_else(
      program_id == "150011", "PAYROLL DEFAULT", costcenter_name
    ),
  ) |>
  select(!c(row_485, is_school))

personnel |> nrow()
```

## Manual fixes

```{r}
positions_fix <-
  read_excel(
    glue("{path_data}/manual/positions_fix.xlsx"),
    col_types = "text"
  ) |>
  distinct(position_id, position_title)

positions_fix |>
  count(position_id, sort = TRUE) |>
  filter(n > 1)
```

# Tidy

## Standardized names

```{r}
program_names <-
  personnel |>
  distinct(program_id, program_name) |>
  filter_out(program_id %in% ledger_full$program_id) |>
  bind_rows(
    ledger_full |> distinct(program_id, program_name),
  )

costcenter_names <-
  personnel |>
  distinct(costcenter_id, costcenter_name) |>
  filter_out(costcenter_id %in% ledger_full$costcenter_id) |>
  bind_rows(
    ledger_full |> distinct(costcenter_id, costcenter_name),
  )
```

Make sure that each unique ID corresponds to only one lookup name for:
* Program name
* Cost center name
* Position title
* Employee name

Use lookup names from ledger, rather than personnel.

```{r}
position_titles <-
  personnel |>
  filter(!is.na(position_title)) |>
  distinct(position_id, position_title, pay_end_dt) |>
  slice_max(order_by = pay_end_dt, n = 1, by = position_id) |>
  select(position_id, position_title)

position_titles <-
  positions_fix |>
  filter(!(position_id %in% position_titles$position_id)) |>
  bind_rows(position_titles)

employee_names <-
  personnel |>
  filter(!is.na(employee_name)) |>
  distinct(employee_id, employee_name, pay_end_dt) |>
  slice_max(order_by = pay_end_dt, n = 1, by = employee_id) |>
  select(employee_id, employee_name)
```

```{r}
personnel <-
  personnel |>
  select(!c(
    program_name,
    costcenter_name,
    position_title,
    employee_name
  )) |>
  left_join(program_names, by = join_by(program_id)) |>
  left_join(costcenter_names, by = join_by(costcenter_id)) |>
  left_join(position_titles, by = join_by(position_id)) |>
  left_join(employee_names, by = join_by(employee_id))
```

## Missing cost centers

Find more accurate cost centers for records with data quality issues.

```{r}
filter_is_bad <- function(df) {
  df |>
    dplyr::filter(
      dplyr::when_any(
        .data$program_id == "150011", # payroll default
        .data$costcenter_id == "40232", # federal programs
        .data$costcenter_id %in% costcenters_negative$costcenter_id,
      )
    )
}

filter_not_bad <- function(df) {
  df |>
    dplyr::filter_out(
      dplyr::when_any(
        .data$program_id == "150011", # payroll default
        .data$costcenter_id == "40232", # federal programs
        .data$costcenter_id %in% costcenters_negative$costcenter_id,
      )
    )
}
```

```{r}
has_program_position <-
  personnel |>
  filter_not_bad() |>
  slice_max(
    order_by = pay_end_dt, n = 1, by = position_id, with_ties = FALSE
  ) |>
  select(position_id, program_id, program_name, costcenter_id, costcenter_name)

has_program_employee <-
  personnel |>
  filter_not_bad() |>
  slice_max(
    order_by = pay_end_dt, n = 1, by = employee_id, with_ties = FALSE
  ) |>
  select(employee_id, program_id, program_name, costcenter_id, costcenter_name)

program_foundby_position <-
  personnel |>
  filter_is_bad() |>
  filter(position_id %in% has_program_position$position_id) |>
  distinct(position_id) |>
  left_join(has_program_position, by = join_by(position_id))

program_foundby_employee <-
  personnel |>
  filter(!(position_id %in% program_foundby_position$position_id)) |>
  filter_is_bad() |>
  filter(employee_id %in% has_program_employee$employee_id) |>
  distinct(employee_id) |>
  left_join(has_program_employee, by = join_by(employee_id))

personnel_foundby_position <-
  personnel |>
  filter_is_bad() |>
  filter(position_id %in% program_foundby_position$position_id) |>
  select(!c(program_id, program_name, costcenter_id, costcenter_name)) |>
  left_join(program_foundby_position, by = join_by(position_id))

personnel_foundby_employee <-
  personnel |>
  filter(!(position_id %in% program_foundby_position$position_id)) |>
  filter_is_bad() |>
  filter(employee_id %in% program_foundby_employee$employee_id) |>
  select(!c(program_id, program_name, costcenter_id, costcenter_name)) |>
  left_join(program_foundby_employee, by = join_by(employee_id))

personnel_notfound <-
  personnel |>
  filter_is_bad() |>
  filter(!(position_id %in% program_foundby_position$position_id)) |>
  filter(!(employee_id %in% program_foundby_employee$employee_id))

personnel <-
  personnel |>
  filter_not_bad() |>
  bind_rows(
    personnel_foundby_position,
    personnel_foundby_employee,
    personnel_notfound
  )
```

```{r}
position_has_school <-
  personnel |>
  filter(costcenter_id == "SCHOOL") |>
  distinct(position_id) |>
  mutate(position_has_school = TRUE)
```

```{r}
personnel <-
  personnel |>
  mutate(
    position_is_direct = str_detect(
      position_title, regex(positions_direct, ignore_case = TRUE)
    )
  ) |>
  left_join(position_has_school, by = join_by(position_id)) |>
  mutate(
    costcenter_id = if_else(
      position_is_direct & position_has_school, "SCHOOL",
      costcenter_id, missing = costcenter_id
    ),
    costcenter_name = if_else(
      position_is_direct & position_has_school, "SCHOOL",
      costcenter_name, missing = costcenter_name
    ),
  )

personnel |> nrow()
```

## Alignment to ledger

Align personnel amounts to ledger amounts.

Calculate multiplier so that personnel expenses equal ledger expenses.

```{r}
ledger_costcenter_account <-
  ledger_personnel |>
  summarize(
    amount = sum(ending_balance),
    .by = c(costcenter_id, costcenter_name, account_id)
  ) |>
  filter(amount != 0.00)

addtnl_rows_costcenter <-
  ledger_costcenter_account |>
  filter_out(costcenter_id %in% personnel$costcenter_id) |>
  mutate(
    program_id = "000000", program_name = "NO PROGRAM",
    .before = costcenter_id
  ) |>
  mutate(
    position_id = "00000000", position_title = "NO POSITION",
    employee_id = "00000000", employee_name = "NO EMPLOYEE",
    .after = account_id
  )

personnel <-
  personnel |>
  bind_rows(addtnl_rows_costcenter) |>
  mutate(account_4 = str_extract(account_id, "^701[1-9]")) |>
  filter(
    costcenter_id %in% costcenters_nonzero$costcenter_id,
    account_4 %in% accounts_4_nonzero$account_4
  )

personnel_account_4 <-
  personnel |>
  summarize(amount = sum(amount), .by = account_4)
```

```{r}
multipliers_4 <-
  accounts_4_nonzero |>
  rename(amount_ledger = ending_balance) |>
  full_join(
    personnel_account_4,
    by = join_by(account_4)
  ) |>
  filter(!is.na(amount)) |>
  mutate(multiplier = amount_ledger / amount) |>
  select(account_4, multiplier)

personnel_aligned <-
  personnel |>
  left_join(multipliers_4, by = join_by(account_4)) |>
  mutate(amount = amount * multiplier) |>
  select(!multiplier) |>
  filter(amount != 0.00)

ledger_personnel |>
  pull(ending_balance) |>
  sum()

personnel_aligned |>
  pull(amount) |>
  sum()
```

# Transform

Roll up to cost center and program.
Add columns for salary, benefits, and terminal leave.

```{r}
personnel_clean <-
  personnel_aligned |>
  mutate(amt_salary = case_when(
    str_starts(account_id, "7011") ~ amount,
    str_starts(account_id, "7012") ~ amount,
    str_starts(account_id, "7013") ~ amount,
    str_starts(account_id, "7015") ~ amount,
    .default = 0.00
  )) |>
  mutate(amt_benefits = case_when(
    str_starts(account_id, "7014") ~ amount,
    .default = 0.00
  )) |>
  mutate(amt_total = amt_salary + amt_benefits) |>
  mutate(amt_terminal = case_when(
    account_id == "7013007" ~ amount,
    .default = 0.00
  )) |>
  summarize(
    amt_salary = sum(amt_salary),
    amt_benefits = sum(amt_benefits),
    amt_total = sum(amt_total),
    amt_terminal = sum(amt_terminal),
    .by = c(
      position_has_school, position_is_direct,
      costcenter_id, costcenter_name,
      program_id, program_name,
      position_id, position_title,
      employee_id, employee_name
    )
  ) |>
  left_join(costcenters_categorization, by = join_by(costcenter_id)) |>
  relocate(c(position_is_direct), .after = last_col()) |>
  arrange(
    is_school, is_all_direct, is_all_indirect, is_mixed,
    costcenter_name,
    program_name,
    position_is_direct, position_title, position_has_school,
    employee_name
  ) |>
  filter(amt_total != 0.00)
```

# Output

```{r}
personnel_clean |>
  write_parquet(glue("{path_data}/working/personnel_clean.parquet"))

personnel_clean |>
  write_csv(glue("{path_data}/working/personnel_clean.csv"), na = "")

personnel_clean |>
  filter(when_any(is_all_direct, is_all_indirect)) |>
  write_csv(glue("{path_data}/working/personnel_rev_none.csv"), na = "")

personnel_clean |>
  filter_out(when_any(is_all_direct, is_all_indirect)) |>
  filter(position_is_direct) |>
  write_csv(glue("{path_data}/working/personnel_rev_direct.csv"), na = "")

personnel_clean |>
  filter_out(when_any(is_all_direct, is_all_indirect)) |>
  filter(!position_is_direct) |>
  write_csv(glue("{path_data}/working/personnel_rev_full.csv"), na = "")
```

# Diagnostics

```{r}
personnel_bad <-
  personnel_clean |>
  mutate(
    bad_program = is.na(program_name),
    bad_costcenter = is.na(costcenter_name),
    bad_position = is.na(position_title),
    bad_employee = is.na(employee_name)
  ) |>
  filter(
    when_any(
      bad_program, bad_costcenter, bad_position, bad_employee,
    )
  ) |>
  group_by(
    bad_program, bad_costcenter, bad_position, bad_employee,
    program_name, program_id,
    costcenter_name, costcenter_id,
    position_title, position_id,
    employee_name, employee_id
  ) |>
  summarize(
    amt_total = sum(amt_total), .groups = "drop"
  )

personnel_bad |>
  filter(bad_program) |>
  distinct(
    position_id, position_title, employee_id, employee_name,
    program_id, program_name, costcenter_id, costcenter_name,
  ) |>
  write_csv(glue("{path_data}/bad/bad_program.csv"), na = "")

personnel_bad |>
  filter(bad_costcenter) |>
  distinct(
    position_id, position_title, employee_id, employee_name,
    program_id, program_name, costcenter_id, costcenter_name,
  ) |>
  write_csv(glue("{path_data}/bad/bad_costcenter.csv"), na = "")

personnel_bad |>
  filter(bad_position) |>
  distinct(
    position_id, position_title, employee_id, employee_name,
    program_id, program_name, costcenter_id, costcenter_name,
  ) |>
  write_csv(glue("{path_data}/bad/bad_position.csv"), na = "")

personnel_bad |>
  filter(bad_employee) |>
  distinct(
    position_id, position_title, employee_id, employee_name,
    program_id, program_name, costcenter_id, costcenter_name,
  ) |>
  write_csv(glue("{path_data}/bad/bad_employee.csv"), na = "")
```

All direct

```{r}
personnel_clean |>
  filter(when_any(is_all_direct)) |>
  summarize(
    .by = c(costcenter_name, costcenter_id),
    n_total = n(),
    n_likely_direct = sum(
      position_has_school | position_is_direct,
      na.rm = TRUE
    ),
    n_school = sum(position_has_school, na.rm = TRUE),
    n_direct = sum(position_is_direct),
    n_indirect = sum(!position_is_direct),
    prop_likely_direct = n_likely_direct / n_total,
    prop_direct = n_direct / n_total
  ) |>
  arrange(costcenter_name)
```

All indirect

```{r}
personnel_clean |>
  filter(when_any(is_all_indirect)) |>
  summarize(
    .by = c(costcenter_name, costcenter_id),
    n_total = n(),
    n_likely_direct = sum(
      position_has_school | position_is_direct,
      na.rm = TRUE
    ),
    n_school = sum(position_has_school, na.rm = TRUE),
    n_direct = sum(position_is_direct),
    n_indirect = sum(!position_is_direct),
    prop_likely_direct = n_likely_direct / n_total,
    prop_direct = n_direct / n_total
  ) |>
  arrange(costcenter_name)
```

Mixed

```{r}
personnel_clean |>
  filter_out(when_any(is_all_direct, is_all_indirect)) |>
  summarize(
    .by = costcenter_name,
    n_total = n(),
    n_likely_direct = sum(
      position_has_school | position_is_direct,
      na.rm = TRUE
    ),
    n_school = sum(position_has_school, na.rm = TRUE),
    n_direct = sum(position_is_direct),
    n_indirect = sum(!position_is_direct),
    prop_likely_direct = n_likely_direct / n_total,
    prop_direct = n_direct / n_total
  ) |>
  arrange(costcenter_name)
```

```{r}
Sys.time()
```
