---
title: "Clean Personnel"
format: html
---

# Setup

```{r}
library(tidyverse)
library(glue)
library(janitor)
library(readxl)
library(writexl)
library(arrow)

options(scipen = 12, digits = 12)

path_data <- Sys.getenv("PATH_DATA")
year_fiscal <- Sys.getenv("YEAR_FISCAL")
```

# Import

Lookup tables

```{r}
#| eval: false

programs <- read_excel("data/programs.xlsx")

costcenters <-
  read_excel(glue("{path_working}/Cost Centers.xlsx"), col_types = "text") %>%
  clean_names() %>%
  select(!costcenter_name)

categorization_prev <-
  read_excel("data/personnel_2023.xlsx", col_types = "text") %>%
  clean_names() %>%
  mutate(
    across(position, \(x) str_pad(x, width = 8, side = "left", pad = "0"))
  ) %>%
  distinct(costcenter, position, categorization_prev = position_type) %>%
  arrange(costcenter, position, categorization_prev)

categorization_sp <-
  read_excel(
    glue(
      "{path_working}/Categorization by Cost Center, Program 2, Program.xlsx"
    ),
    col_types = "text"
  ) %>%
  clean_names() %>%
  select(costcenter, program_2, program, categorization_sp)

categorization_override <-
  read_excel(
    glue(
      "{path_working}/Override Personnel.xlsx"
    ),
    col_types = "text"
  ) %>%
  clean_names() %>%
  filter(!is.na(categorization_override)) %>%
  distinct(costcenter, position, categorization_override)

headcomponent <-
  read_excel(
    glue(
      "{path_working}/Head of Component.xlsx"
    ),
    col_types = "text"
  ) %>%
  clean_names() %>%
  distinct(costcenter, position, is_executive, is_headcomponent)

titles_manual <-
  read_excel(
    glue(
      "{path_working}/Missing - Title.xlsx"
    ),
    col_types = "text"
  ) %>%
  clean_names() %>%
  distinct(position, title_manual)

payroll_default <-
  read_excel(
    glue(
      "{path_working}/Payroll Default.xlsx"
    ),
    col_types = "text"
  ) %>%
  clean_names() %>%
  select(!note) %>%
  mutate(is_default = as.logical(is_default))
```

Ledger

```{r}
ledger <-
  read_parquet(glue("{path_data}/ledger.parquet")) %>%
  filter(
    account_category == "700000A",
    account_id != "7012999"
  ) %>%
  filter(ending_balance != 0.00)

ledger_rollup <-
  ledger %>%
  summarize(expense_ledger = sum(ending_balance), .by = c(
    account_id, account_name,
  )) %>%
  arrange(account_id) %>%
  filter(expense_ledger != 0.00)
```

Personnel

```{r}
data_485_raw <-
  read_parquet(glue("{path_data}/personnel_{year_fiscal}.parquet")) %>%
  replace_na(list(project_id = "000000", award_id = "0000000")) %>%
  select(!c(row_485, account_raw, account_pref_2, account_pref_4))

data_485 <-
  data_485_raw %>%
  filter(costcenter_id %in% ledger$costcenter_id) %>%
  filter(
    account_id != "spc", # not an account ID
    account_id != "3111003", # not personnel
    account_id != "7013002", # not in ledger
    account_id != "7016001" # not in ledger
  )
```

# Tidy

Initial cleaning

```{r}
program_names <-
  ledger %>%
  distinct(program_id, program_name)

costcenter_names <-
  ledger %>%
  distinct(costcenter_id, costcenter_name)

project_names <-
  ledger %>%
  distinct(project_id, project_name)

award_names <-
  ledger %>%
  distinct(award_id, award_name)

position_titles <-
  data_485_raw %>%
  distinct(position_id, position_title, pay_end_dt) %>%
  slice_max(order_by = pay_end_dt, n = 1, by = position_id) %>%
  select(position_id, position_title)

employee_names <-
  data_485_raw %>%
  distinct(employee_id, employee_name, pay_end_dt) %>%
  slice_max(order_by = pay_end_dt, n = 1, by = employee_id) %>%
  select(employee_id, employee_name)
```

```{r}
data_485 <-
  data_485 %>%
  select(!c(
    program_name,
    costcenter_name,
    project_name,
    award_name,
    position_title,
    employee_name
  )) %>%
  left_join(program_names, by = join_by(program_id)) %>%
  left_join(costcenter_names, by = join_by(costcenter_id)) %>%
  left_join(project_names, by = join_by(project_id)) %>%
  left_join(award_names, by = join_by(award_id)) %>%
  left_join(position_titles, by = join_by(position_id)) %>%
  left_join(employee_names, by = join_by(employee_id))
```

Compare accounts in ledger to accounts in 485.

Find ledger accounts missing from 485.
For these accounts, additional rows will be imputed into the 485.

Find accounts present in ledger and in 485.
For these accounts, a multiplier will calculated per account so that personnel 
spending in the 485 equals personnel spending in the ledger.

```{r}
personnel_rollup <-
  data_485 %>%
  summarize(expense_485 = sum(amount), .by = c(
    account_id
  )) %>%
  arrange(account_id) %>%
  filter(expense_485 != 0.00)

personnel_comparison <-
  ledger_rollup %>%
  full_join(personnel_rollup, by = join_by(
    account_id
  ))

accounts_missing_485 <-
  personnel_comparison %>%
  filter(is.na(expense_485)) %>%
  select(account_id, expense_ledger)

multipliers <-
  personnel_comparison %>%
  filter(!is.na(expense_485)) %>%
  mutate(multiplier = expense_ledger / expense_485) %>%
  select(account_id, multiplier)
```

Impute

```{r}
attributes_imputable <-
  ledger %>%
  filter(account_id %in% accounts_missing_485$account_id) %>%
  distinct(
    fund_id, program_id, costcenter_id,
    account_id
  ) %>%
  mutate(is_imputed = TRUE) %>%
  pivot_wider(
    names_from = account_id, names_sort = TRUE,
    values_from = is_imputed
  )

rows_need_impute <-
  data_485 %>%
  left_join(attributes_imputable, by = join_by(
    fund_id, program_id, costcenter_id,
  )) %>%
  filter(if_any(`7013015`:`7014021`)) %>%
  select(!c(amount, account_id)) %>%
  pivot_longer(
    cols = c(`7013015`:`7014021`),
    names_to = "account_id", values_to = "is_imputed", values_drop_na = TRUE
  ) %>%
  select(!is_imputed)

amounts_impute <-
  rows_need_impute %>%
  count(account_id) %>%
  left_join(accounts_missing_485, by = join_by(account_id)) %>%
  mutate(amount = expense_ledger / n) %>%
  select(account_id, amount) %>%
  arrange(account_id)

rows_imputed <-
  rows_need_impute %>%
  left_join(amounts_impute, by = join_by(account_id))
```

```{r}
data_485_corrected <-
  data_485 %>%
  left_join(multipliers, by = join_by(account_id)) %>%
  mutate(amount = amount * multiplier) %>%
  select(!multiplier) %>%
  bind_rows(rows_imputed)

data_485_corrected %>%
  pull(amount) %>%
  sum()

ledger %>%
  pull(ending_balance) %>%
  sum()
```

Find correct programs for payroll default records.

```{r}
has_program_position <-
  data_485_corrected %>%
  filter(!is.na(program_name)) %>%
  slice_max(
    order_by = pay_end_dt, n = 1, by = position_id, with_ties = FALSE
  ) %>%
  select(position_id, program_id, program_name)

has_program_employee <-
  data_485_corrected %>%
  filter(!is.na(program_name)) %>%
  slice_max(
    order_by = pay_end_dt, n = 1, by = employee_id, with_ties = FALSE
  ) %>%
  select(employee_id, program_id, program_name)

program_foundby_position <-
  data_485_corrected %>%
  filter(is.na(program_name)) %>%
  distinct(position_id) %>%
  filter(position_id %in% has_program_position$position_id) %>%
  left_join(has_program_position, by = join_by(position_id)) %>%
  arrange(program_name, position_id)

program_foundby_employee <-
  data_485_corrected %>%
  filter(is.na(program_name)) %>%
  filter(!(position_id %in% program_foundby_position$position_id)) %>%
  distinct(employee_id) %>%
  filter(employee_id %in% has_program_employee$employee_id) %>%
  left_join(has_program_employee, by = join_by(employee_id)) %>%
  arrange(program_name, employee_id)

data_485_corrected %>%
  filter(is.na(program_name)) %>%
  distinct(costcenter_name, program_id, position_id) %>%
  count(costcenter_name, program_id)
```

Find correct cost centers for payroll default records.

```{r}
has_costcenter_position <-
  data_485_corrected %>%
  filter(program_id != "150011") %>%
  slice_max(
    order_by = pay_end_dt, n = 1, by = position_id, with_ties = FALSE
  ) %>%
  select(position_id, costcenter_id, costcenter_name)

has_costcenter_employee <-
  data_485_corrected %>%
  filter(program_id != "150011") %>%
  slice_max(
    order_by = pay_end_dt, n = 1, by = employee_id, with_ties = FALSE
  ) %>%
  select(employee_id, costcenter_id, costcenter_name)

costcenter_foundby_position <-
  data_485_corrected %>%
  filter(program_id == "150011") %>%
  distinct(position_id) %>%
  filter(position_id %in% has_costcenter_position$position_id) %>%
  left_join(has_costcenter_position, by = join_by(position_id)) %>%
  arrange(costcenter_name, position_id)

costcenter_foundby_employee <-
  data_485_corrected %>%
  filter(program_id == "150011") %>%
  filter(!(position_id %in% costcenter_foundby_position$position_id)) %>%
  distinct(employee_id) %>%
  filter(employee_id %in% has_costcenter_employee$employee_id) %>%
  left_join(has_costcenter_employee, by = join_by(employee_id)) %>%
  arrange(costcenter_name, employee_id)
```

***

# Transform

```{r}
data_485_clean <-
  data_485_corrected %>%
  mutate(
    amt_pay_base = rowSums(across(starts_with("amt_7011") | starts_with("amt_7012"))),
    amt_pay_additonal = rowSums(across(starts_with("amt_7013"))),
    amt_overtime = rowSums(across(starts_with("amt_7015"))),
    amt_exception = rowSums(across(starts_with("amt_7016"))),
  ) %>%
  mutate(
    amt_salary = amt_pay_base + amt_pay_additonal + amt_overtime + amt_exception,
    amt_benefits = rowSums(across(starts_with("amt_7014"))),
    amt_terminal = amt_7013007
  )
```



```{r}
costcenters_default <-
  data_485_clean %>%
  filter(program_name == "PAYROLL DEFAULT") %>%
  distinct(emplid) %>%
  left_join(data_485_clean, by = join_by(emplid)) %>%
  filter(program_name != "PAYROLL DEFAULT") %>%
  group_by(emplid, costcenter, costcenter_name, program, program_name) %>%
  summarize(amt_total = sum(amt_salary + amt_benefits)) %>%
  group_by(emplid) %>%
  slice_max(order_by = amt_total, n = 1) %>%
  ungroup() %>%
  select(
    emplid,
    costcenter_default = costcenter, costcenter_name_default = costcenter_name,
    program_default = program, program_name_default = program_name
  ) %>%
  mutate(is_default = TRUE)

payroll_default_dedupe <-
  payroll_default %>%
  filter(!(emplid %in% costcenters_default$emplid)) %>%
  bind_rows(costcenters_default)
```

Augment

```{r}
data_485_aug <-
  data_485_clean %>%
  mutate(is_default = if_else(program_name == "PAYROLL DEFAULT", TRUE, FALSE)) %>%
  left_join(payroll_default_dedupe, by = join_by(emplid, is_default)) %>%
  mutate(
    costcenter = if_else(!is.na(costcenter_default), costcenter_default, costcenter),
    costcenter_name = if_else(!is.na(costcenter_name_default), costcenter_name_default, costcenter_name),
    program = if_else(!is.na(program_default), program_default, program),
    program_name = if_else(!is.na(program_name_default), program_name_default, program_name)
  ) %>%
  left_join(costcenters, by = join_by(costcenter)) %>%
  left_join(programs, by = join_by(program)) %>%
  left_join(pfr_start, by = join_by(position)) %>%
  left_join(pfr_end, by = join_by(position)) %>%
  left_join(titles_manual, by = join_by(position)) %>%
  left_join(names_clean, by = join_by(emplid)) %>%
  left_join(headcomponent, by = join_by(costcenter, position)) %>%
  left_join(categorization_override, by = join_by(costcenter, position)) %>%
  left_join(categorization_prev, by = join_by(costcenter, position)) %>%
  left_join(categorization_sp, by = join_by(costcenter, program_2, program)) %>%
  mutate(title = case_when(
    !is.na(title_end) ~ title_end,
    !is.na(title_start) ~ title_start,
    .default = title_manual
  )) %>%
  mutate(headcomponent = case_when(
    is_executive == 1 ~ "Executive",
    is_headcomponent == 1 ~ "Head of component",
    .default = "Non-executive"
  )) %>%
  mutate(
    categorization = case_when(
      costcenter_name_clean == "SCHOOLS" ~ "Direct",
      program_name == "PAYROLL DEFAULT" ~ "Indirect",
      !is.na(categorization_override) ~ categorization_override,
      !is.na(categorization_prev) ~ categorization_prev,
      .default = categorization_sp
    )
  ) %>%
  mutate(
    costcenter_clean = if_else(program_name == "PAYROLL DEFAULT", "00000", costcenter_clean),
    costcenter_name_clean = if_else(program_name == "PAYROLL DEFAULT", "NO COST CENTER", costcenter_name_clean)
  )
```

# Output

```{r}
personnel <-
  data_485_aug %>%
  group_by(
    costcenter_clean, costcenter, program_2, program, position, emplid, headcomponent,
    costcenter_name_clean, costcenter_name, program_2_name, program_name, title, name_clean,
    categorization
  ) %>%
  summarize(
    amt_salary = sum(amt_salary),
    amt_benefits = sum(amt_benefits),
    amt_terminal = sum(amt_terminal)
  ) %>%
  ungroup() %>%
  relocate(
    c(categorization, costcenter_name_clean, costcenter_name, program_2_name, program_name, title, name_clean, headcomponent)
  ) %>%
  mutate(
    amt_direct = if_else(categorization == "Direct", amt_salary + amt_benefits, 0),
    amt_indirect = if_else(categorization == "Indirect", amt_salary + amt_benefits, 0),
    amt_total = amt_direct + amt_indirect,
    amt_unrestricted = if_else(headcomponent == "Non-executive", amt_salary + amt_benefits, 0.00),
    amt_headcomponent_direct = if_else(headcomponent != "Non-executive" & categorization == "Direct", amt_total, 0),
    amt_headcomponent_indirect = if_else(headcomponent != "Non-executive" & categorization == "Indirect", amt_total, 0),
    amt_headcomponent_total = if_else(headcomponent != "Non-executive", amt_total, 0),
    amt_terminal_to_indirect = if_else(categorization == "Direct" & headcomponent == "Non-executive", amt_terminal, 0),
    amt_terminal_restricted = if_else(headcomponent != "Non-executive", amt_terminal, 0)
  ) %>%
  arrange(costcenter_name_clean, costcenter_name, program_2_name, program_name, title, name_clean) %>%
  relocate(c(costcenter_clean, costcenter, program_2, program, position, emplid), .after = last_col())

personnel %>% write_csv("data/csv/personnel.csv", na = "")
```

# Diagnostics

Unique combinations

```{r}
data_485_aug %>%
  distinct(
    costcenter_name_clean, costcenter_name, program_name, title, name_clean,
    costcenter_clean, costcenter, program_2, position, emplid,
    categorization
  ) %>%
  nrow()
```

Every position is associated with only 1 title.

PASS

```{r}
data_485_aug %>%
  distinct(position, title) %>%
  count(position, sort = T) %>%
  filter(n > 1)
```

Every employee ID is associated with only 1 name.

PASS

```{r}
data_485_aug %>%
  distinct(emplid, name_clean) %>%
  count(emplid, sort = T) %>%
  filter(n > 1)
```

Every cost center, position group is associated with only 1 categorization.

PASS

```{r}
data_485_aug %>%
  distinct(costcenter, position, categorization) %>%
  count(costcenter, position, sort = T) %>%
  filter(n > 1)
```

Cost centers with only 1 categorization.

```{r}
personnel %>%
  distinct(costcenter_name_clean, categorization) %>%
  count(costcenter_name_clean, sort = TRUE) %>%
  filter(n == 1) %>%
  select(costcenter_name_clean) %>%
  left_join(personnel) %>%
  count(costcenter_name_clean, costcenter_clean, categorization)
```

Cost centers with more than 1 categorization.

```{r}
personnel %>%
  filter(costcenter != "10002") %>%
  distinct(costcenter_name_clean, categorization) %>%
  count(costcenter_name_clean, sort = TRUE) %>%
  filter(n > 1) %>%
  select(costcenter_name_clean) %>%
  left_join(personnel) %>%
  count(costcenter_name_clean, costcenter_clean, categorization)
```

```{r}
personnel %>%
  filter(program_name == "PAYROLL DEFAULT")
```

 $1,168,754,375.06 

```{r}
personnel %>%
  pull(amt_total) %>%
  sum()
```

```{r}
Sys.time()
```
